---
title: "Housing Analysis with R"
author: "Greg Cooke"
date: "12/05/2020"
output: 
  bookdown::html_document2
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

```

# Introduction and Questions I want to answer{-}

In this project I am looking at the UK's housing economy, I have chosen this topic because it is an aim of mine to be a homeowner and it is a topic which obviously strikes fear into a lot of millenials. There are questions I want to answer:

- What areas have the highest growth over the time period?
- What are the biggest and the most common factors in the price of a house?
- How does owning a house compare as an investment?

As with any projects I load in my packages, there's quite a few I've used here and some were for simple experimentation so I will remove them when necersary.

```{r Libraries, results = "hide",message = FALSE,warning = FALSE}

library(tidyverse)
library(lubridate)
library(mgsub)
library(stringi)
library(class)
library(data.table)
library(RSQLite)
library(DBI)
library(gridExtra)
library(openxlsx)
library(broom)
library(ggraph)
library(ggeffects)
library(class)

```

This is followed by cleaning and tidying the datasets with dplyr and tidyr, I have created one big data frame from this called hpi_lad.

```{r Setup, results = "hide",message = FALSE,warning = FALSE}

## Load and Tidy RPI Dataset

hpi <- read_csv(paste0(getwd(),'/HPI_Full.csv'), col_names = T) %>%
  dplyr:: mutate(Date = as_date(dmy(Date))) %>%
  dplyr:: select(-IndexSA)

## Load and Tidy 2011 Census Dataset

census2011_lad <- read.xlsx(paste0(getwd(),'/censusdata.xlsx'), sheet = 3, startRow = 8) %>%
  dplyr:: rename(AreaCode = Code,
                 RegionCounty = `Region/Country`,
                 Supergroup = Supergroup.name,
                 Group = Group.name,
                 Subgroup = Subgroup.name) %>%
  dplyr:: select(-c(Name, contains(".code"), `Country/region`)) %>%
  dplyr:: mutate_if(is.character, trimws)

## Load and Tidy Bankrate Dataset

br <- read_csv(paste0(getwd(),'/bankrate.csv'), col_names = T) %>%
  dplyr:: rename(Date = `Date Changed`) %>%
  dplyr:: mutate(Date = as_date(dmy(Date))) %>%
  tidyr:: complete(Date = seq.Date(min(Date), max(Date), by="day")) %>%
  tidyr:: fill(Rate)

# Combine Datasets

hpi_lad <- hpi %>%
  dplyr:: left_join(census2011_lad[ , (names(census2011_lad)) %in% 
                                         c("Name","AreaCode", "RegionCounty", "Supergroup", "Group", "Subgroup")], 
                    by = "AreaCode") %>%
  dplyr:: mutate(Country = case_when(RegionCounty == "Scotland" ~ "Scotland",
                                     RegionCounty == "Wales" ~ "Wales",
                                     RegionCounty == "Northern Ireland" ~ "NorthernIreland",
                                     TRUE ~ "England"),
                 RegionCounty = case_when(RegionCounty == Country ~ RegionName,
                                    RegionCounty != Country ~ RegionCounty)) %>%
  dplyr:: left_join(br, 
                    by = "Date") %>%
  dplyr:: select(Date, AreaCode, RegionName, RegionCounty, Country, 
                 Supergroup, Group, Subgroup, everything()) %>%
  dplyr:: filter(!is.na(RegionCounty)) %>% # this data set has everything but LAD level removed
  dplyr:: rename(Area = RegionName,
                 Region = RegionCounty)

housingtypes <- c("Detached","SemiDetached","Terraced","Flat")
  
rm(hpi,br,census2011_lad)
gc()

```

# What areas have the highest growth over the time period

I created a function to graph how individual areascompare so I could start to understand the data to help me build a plan for how I was going to create an answer for this.

```{r Graph Function 1, message = FALSE, warning = FALSE}

AreaRegionCountryCompareR <- function(df, 
                                      area,
                                      datefil){
  
  reg <- unique(df[df$Area == area,]$Region)
  country <- unique(df[df$Area == area,]$Country)
  places <- c(area, 
              reg, 
              country)
  
  type <- c("Area",
            "Region",
            "Country")
  
  dfcreate <- function(x){
    
    dfgg <- x %>%
      dplyr:: filter_at(vars(contains(type[i])),  any_vars(. == places[i])) %>%
      dplyr:: group_by_at(c(type[i], "Date")) %>%
      dplyr:: summarise(MeanPrice = mean(AveragePrice, na.rm = T)) %>%
      dplyr:: ungroup()
    
    names(dfgg)[1] <- "Location"
    
    return(dfgg)
    
  }
  
  prices <- list()
  
  for (i in 1:length(places)){
    
   prices[[i]] <- assign(paste0(places[i], "Price"), dfcreate(hpi_lad))
    
  }

  dfgg1 <- data.frame(df[df$Area == area,]) %>%
    tidyr:: gather(key = !!eval(expr(paste(!!!area))), 
                   value = 'Price', 
                   all_of(c("AveragePrice",
                            "DetachedPrice",
                            "FlatPrice",
                            "SemiDetachedPrice",
                            "TerracedPrice"))) %>%
    dplyr:: filter(!is.na(Price)|
                   Date >= datefil) %>%
    dplyr:: group_by_at(c(area, "Date")) %>%
    dplyr:: summarise(MeanPrice = mean(Price, na.rm = T),
                      SDPrice = sd(Price, na.rm = T),
                      MeanVol = mean(SalesVolume, na.rm = T),
                      SDVol = sd(SalesVolume, na.rm = T))  

  grid.arrange(
    
    ggplot(bind_rows(prices),
           aes(Date, MeanPrice, colour = Location)) +
      geom_line() +
      ggtitle(paste0(area," Price Comparison")),
    
    ggplot(dfgg1, 
           aes_string(x = "Date", y = "MeanPrice", colour = paste0("`",area,"`")), na.rm = T) +
      geom_line() +
      ggtitle(paste0(area," Housing Price Split")),
    
    nrow = 2)
     
} 

AreaRegionCountryCompareR(df = hpi_lad, 
                          area = "West Berkshire",
                          datefil = "1995-01-01")

```



##### Data References {-}

- [HPI data](https://www.gov.uk/government/statistical-data-sets/uk-house-price-index-data-downloads-february-2020)
- [Bank Rate data](https://www.bankofengland.co.uk/boeapps/database/Bank-Rate.asp)
- [2011 Census Data for LADs](https://www.ons.gov.uk/methodology/geography/geographicalproducts/areaclassifications/2011areaclassifications/datasets)
- [Inflation Time Series](https://www.ons.gov.uk/economy/inflationandpriceindices/datasets/consumerpriceindices)

